---
title: "Tidy Tuesday Exercise"
---

## Loading data and initial exploration
```{r, warning=FALSE}
# Load packages
library(tidyverse)
library(tidymodels)
library(GGally)
library(here)
library(knitr)

# set seed 
rngseed <- 1234
set.seed(rngseed)

# Load data
d1 <- readRDS(here("ml-models-exercise", "clean-mavoglurant-data.rds"))

# quick cheak
glimpse(d1)
summary(d1)
```

## Recode RACE and create BMI
```{r}

d1 <- d1 %>%
  mutate(
    BMI = WT / (HT^2),
    RACE = as.character(RACE),
    RACE = if_else(RACE %in% c("7", "88"), "3", RACE),
    RACE = as.factor(RACE)
  )
glimpse(d1)
summary(d1)
```

## Pairwise correlation plot
```{r}
# Select continuous variables for correlation analysis
cont_vars <- d1 %>%
  select(Y, AGE, WT, HT, BMI)

# Create a pairwise correlation plot
ggpairs(cont_vars)
```

## Create recipes and workflows for each model 
```{r}
# Create a recipe for all models
# Outcome is Y, all other variables are predictors
base_recipe <- recipe(Y ~ ., data = d1) %>%
  # Create dummy variables for categorical predictors
  step_dummy(all_nominal_predictors()) %>%
  # Remove predictors with zero variance
  step_zv(all_predictors()) %>%
  # Normalize numeric predictors
  step_normalize(all_numeric_predictors())
```
```{r}
# Specify a linear regression model
lm_spec <- linear_reg() %>%
  set_engine("lm")

# Specify a LASSO regression model with penalty fixed at 0.1
lasso_spec <- linear_reg(
  penalty = 0.1,
  mixture = 1
) %>%
  set_engine("glmnet")

# Specify a random forest model with default tuning settings
# The seed is passed into the engine for reproducibility
rf_spec <- rand_forest() %>%
  set_engine("ranger", seed = rngseed) %>%
  set_mode("regression")
```
```{r}
# Create workflow for the linear model
lm_wf <- workflow() %>%
  add_recipe(base_recipe) %>%
  add_model(lm_spec)

# Create workflow for the LASSO model
lasso_wf <- workflow() %>%
  add_recipe(base_recipe) %>%
  add_model(lasso_spec)

# Create workflow for the random forest model
rf_wf <- workflow() %>%
  add_recipe(base_recipe) %>%
  add_model(rf_spec)
```

##  Fit the models
```{r}
# Fit the linear model to the full dataset
lm_fit <- fit(lm_wf, data = d1)

# Fit the LASSO model to the full dataset
lasso_fit <- fit(lasso_wf, data = d1)

# Fit the random forest model to the full dataset
rf_fit <- fit(rf_wf, data = d1)
```

## Generate predictions and compute RMSE
```{r}
# Generate predictions for the linear model
lm_pred <- predict(lm_fit, new_data = d1) %>%
  bind_cols(d1 %>% select(Y)) %>%
  mutate(model = "Linear Model")

# Compute RMSE for the linear model
lm_rmse <- lm_pred %>%
  rmse(truth = Y, estimate = .pred)

# Generate predictions for the LASSO model
lasso_pred <- predict(lasso_fit, new_data = d1) %>%
  bind_cols(d1 %>% select(Y)) %>%
  mutate(model = "LASSO")

# Compute RMSE for the LASSO model
lasso_rmse <- lasso_pred %>%
  rmse(truth = Y, estimate = .pred)

# Generate predictions for the random forest model
rf_pred <- predict(rf_fit, new_data = d1) %>%
  bind_cols(d1 %>% select(Y)) %>%
  mutate(model = "Random Forest")

# Compute RMSE for the random forest model
rf_rmse <- rf_pred %>%
  rmse(truth = Y, estimate = .pred)

# Combine RMSE results into one table
bind_rows(lm_rmse, lasso_rmse, rf_rmse)
```

## Observed vs predicted plots

```{r}
# Combine predictions from all models
pred_all <- bind_rows(lm_pred, lasso_pred, rf_pred)

# Create observed versus predicted plots
ggplot(pred_all, aes(x = Y, y = .pred)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  facet_wrap(~ model) +
  labs(
    title = "Observed versus Predicted Values",
    x = "Observed Y",
    y = "Predicted Y"
  ) +
  theme_bw()
```

As can be seen from the figure, the prediction points of the random forest are closer to the diagonal, indicating that its prediction is more accurate. While the prediction results of Lasso and the linear model are similar, because the penalty term of Lasso is smaller, it causes its coefficients to be close to those of the linear model.
## Tune the LASSO model 
```{r}
# Create an apparent resample object
# This uses the same data for both training and evaluation
# It is intentionally incorrect for teaching purposes
app_resamples <- apparent(d1)

# Define a penalty grid for LASSO
# The values range from 1e-5 to 1e2 on a log scale
lasso_grid <- tibble(
  penalty = 10^seq(-5, 2, length.out = 50)
)

# Specify a tunable LASSO model
lasso_tune_spec <- linear_reg(
  penalty = tune(),
  mixture = 1
) %>%
  set_engine("glmnet")

# Create a workflow for the tunable LASSO model
lasso_tune_wf <- workflow() %>%
  add_recipe(base_recipe) %>%
  add_model(lasso_tune_spec)

# Tune the LASSO model using apparent resamples
lasso_tune_app <- tune_grid(
  lasso_tune_wf,
  resamples = app_resamples,
  grid = lasso_grid,
  metrics = metric_set(rmse)
)

# Show tuning metrics
collect_metrics(lasso_tune_app)
```

## Tune the random forest model

```{r}
# Specify a tunable random forest model
# We tune mtry and min_n, and fix the number of trees at 300
rf_tune_spec <- rand_forest(
  mtry = tune(),
  min_n = tune(),
  trees = 300
) %>%
  set_engine("ranger", seed = rngseed) %>%
  set_mode("regression")

# Create a workflow for the tunable random forest model
rf_tune_wf <- workflow() %>%
  add_recipe(base_recipe) %>%
  add_model(rf_tune_spec)

# Define a regular tuning grid for random forest
# mtry ranges from 1 to 7
# min_n ranges from 1 to 21
# 7 levels are used for each parameter
rf_grid <- grid_regular(
  mtry(range = c(1L, 7L)),
  min_n(range = c(1L, 21L)),
  levels = 7
)

# Tune the random forest model using apparent resamples
rf_tune_app <- tune_grid(
  rf_tune_wf,
  resamples = app_resamples,
  grid = rf_grid,
  metrics = metric_set(rmse)
)

# Show tuning metrics
collect_metrics(rf_tune_app)
```

## LASSO and Random Forest tuning with cross-validation

```{r}
# Reset the random seed before creating cross-validation folds
set.seed(rngseed)

# Create 5-fold cross-validation repeated 5 times
cv_folds <- vfold_cv(d1, v = 5, repeats = 5)
```


```{r}
# Tune the LASSO model using repeated cross-validation
lasso_tune_cv <- tune_grid(
  lasso_tune_wf,
  resamples = cv_folds,
  grid = lasso_grid,
  metrics = metric_set(rmse)
)

# Plot tuning results from repeated CV
autoplot(lasso_tune_cv)

# Show tuning metrics
collect_metrics(lasso_tune_cv)

# Show the best penalty value based on RMSE
show_best(lasso_tune_cv, metric = "rmse")
```

```{r}
# Tune the random forest model using repeated cross-validation
rf_tune_cv <- tune_grid(
  rf_tune_wf,
  resamples = cv_folds,
  grid = rf_grid,
  metrics = metric_set(rmse)
)

# Plot tuning results from repeated CV
autoplot(rf_tune_cv)

# Show tuning metrics
collect_metrics(rf_tune_cv)

# Show the best tuning parameter combination based on RMSE
show_best(rf_tune_cv, metric = "rmse")
```

Overall, the results show that model flexibility can improve performance, but proper evaluation matters. In the initial full-data fit, the random forest produced predictions closest to the observed values, while the LASSO and linear model performed almost identically. The LASSO tuning results showed that very small penalty values gave the lowest RMSE, which makes sense because a small penalty makes the LASSO behave similarly to ordinary linear regression. As the penalty increased, RMSE rose because stronger regularization shrank the coefficients too much and reduced predictive accuracy. For the random forest, better performance was observed when more features were considered at each split and smaller node sizes were allowed, suggesting that a more flexible tree structure fit the data better. However, the tuning results from apparent resampling were overly optimistic because the same data were used for both training and evaluation, which is why cross-validation is essential for reliable model assessment. The cross-validation results provided more realistic estimates of predictive performance and showed that the random forest remained the best-performing model overall.