---
title: "Data Exercise(Week4)"
---

# Option2: Synthetic Data Generation and Exploartion 

## Generate synthetic Data : 
* In this section, I generate a synthetic dataset (N=1000) representing an older adult population (50+). 
The underlying logic includes:
1.  Age & Depression: Strong risk factors for dementia.
2.  Education & Physical Activity: Protective factors.
3.  Comorbidities (Diabetes/Hypertension): Driven by Age and BMI.
4.  BMI & Dementia: A U-shaped relationship(both underweight and obese increase risk).

```{r, echo=FALSE, message=FALSE}
# 0) Load required packages
library(tidyverse)
library(ggplot2)
# 1) Reproducibility seed
base::set.seed(1)

# 2) functions
inv_logit <- function(x) 1 / (1 + base::exp(-x))
clamp <- function(x, lower, upper) base::pmax(lower, base::pmin(x, upper))

# 3) Sample size
N <- 1000

# 4) Generate synthetic dataset
synthetic_data <- tibble::tibble(
  ID = base::sprintf("%05d", 1:N),

  # Age: 85% from truncated normal-like around 65, 15% from uniform tails
  age = {
    mix <- stats::rbinom(n = N, size = 1, prob = 0.85)
    main <- clamp(stats::rnorm(n = N, mean = 65, sd = 7.5), 50, 95)
    tail <- stats::runif(n = N, min = 50, max = 95)
    base::as.integer(base::round(base::ifelse(mix == 1, main, tail)))
  },

  # Gender: ~52% female
  gender = {
    g <- stats::rbinom(n = N, size = 1, prob = 0.52)
    base::factor(base::ifelse(g == 1, "Female", "Male"), levels = c("Male", "Female"))
  }
) |>
  dplyr::mutate(
    # Education (years): mean ~13 with mild cohort effect (older -> slightly fewer years)
    education_years = {
      edu_base <- stats::rnorm(n = N, mean = 13, sd = 2.7)
      edu <- clamp(edu_base - 0.03 * (age - 65), 0, 22)
      base::as.integer(base::round(edu))
    },

    # BMI: mean ~27 sd ~5, mild age pattern, clamped
    BMI = {
      bmi_base <- stats::rnorm(n = N, mean = 27, sd = 5)
      bmi_val <- clamp(bmi_base + 0.03 * (pmin(age, 70) - 50), 15, 55)
      base::round(bmi_val, 1)
    },

    # Physical activity: ordered factor influenced by education (+) and age (-)
    physical_activity = {
      latent <- 0.12 * (education_years - 12) - 0.03 * (age - 65) + stats::rnorm(n = N)
      pa <- base::cut(
        latent,
        breaks = c(-Inf, -0.4, 0.8, Inf),
        labels = c("Low", "Moderate", "High"),
        right = TRUE
      )
      base::factor(pa, levels = c("Low", "Moderate", "High"), ordered = TRUE)
    },

    # Depression history: ~15–20% with links to age and physical activity
    depression_history = {
      lp <- -1.6 +
        0.012 * (age - 65) +
        0.25  * (physical_activity == "Low") -
        0.10  * (physical_activity == "High") +
        0.08  * (gender == "Female")
      stats::rbinom(n = N, size = 1, prob = inv_logit(lp))
    },

    # Diabetes and hypertension: probabilities increase with age and BMI
    diabetes = {
      lp <- -2.0 + 0.035 * (age - 65) + 0.070 * (BMI - 27) + 0.15 * (gender == "Male")
      stats::rbinom(n = N, size = 1, prob = inv_logit(lp))
    },

    hypertension = {
      lp <- -1.0 + 0.050 * (age - 65) + 0.040 * (BMI - 27) + 0.10 * (gender == "Male")
      stats::rbinom(n = N, size = 1, prob = inv_logit(lp))
    },

    # Number of comorbidities (0,1,2) among diabetes + hypertension
    n_comorbidities = diabetes + hypertension,

    # Dementia (0/1): age & depression strong risk; edu & PA protective;
    # diabetes & hypertension risk; BMI U-shaped via quadratic term
    dementia = {
      bmi_u <- ((BMI - 27)^2) / 25
      lp <- -3.2 +
        0.075 * (age - 65) +
        0.85  * depression_history +
        (-0.08) * (education_years - 13) +
        (-0.25) * (physical_activity == "Moderate") +
        (-0.50) * (physical_activity == "High") +
        0.45  * diabetes +
        0.35  * hypertension +
        0.55  * bmi_u
      stats::rbinom(n = N, size = 1, prob = inv_logit(lp))
    }
  )
```

# Exploration with plots

```{r} 
# Summary statistics
summary(synthetic_data)
str(synthetic_data)
```

## Age distribution 
From the histogram, it can be seen that the age range is mainly concentrated between 55 and 75.
```{r}
ggplot(synthetic_data, aes(age)) +
  geom_histogram(bins = 30) +
  labs(title = "Age distribution", x = "Age", y = "Count")
```

## BMI distribution
From the histogram, it can be seen that the BMI range is mainly concentrated between 25 and 30.
```{r}
ggplot(synthetic_data, aes(BMI)) +
  geom_histogram(bins = 30) +
  labs(title = "BMI distribution", x = "BMI", y = "Count")
```

## Age trend with Dementia
The smooth curve shows the trend of dementia probability increasing with age, which is as expected, as age was deliberately made a significant risk factor in the data generation process.
```{r}
ggplot(synthetic_data, aes(x = age, y = dementia)) +
  geom_smooth(method = "loess", color = "blue") + 
  labs(title = "Dementia Risk increases with Age", y = "Probability")
```

## Check BMI with dementia
The curve indicates a non-linear relationship between BMI and the risk of dementia. Its shape conforms to the expected U-shaped association, where the risk increases both when BMI is low and when it is high.
```{r}
ggplot(synthetic_data, aes(x = BMI, y = dementia)) +
  geom_smooth(method = "loess") +
  labs(title = "U-shaped BMI relationship with dementia risk", x = "BMI", y = "Probability (loess-smoothed)")
```

# Fit model 
## Model 1: logistic regression 
From the results,
```{r}
m1 <- glm(
  dementia ~ age + depression_history + education_years + physical_activity +
    diabetes + hypertension + BMI,
  data = synthetic_data,
  family = binomial()
)
summary(m1)
exp(coef(m1))
```

## # Model 2: logistic regression + bmi^2
```{r}
m2 <- glm(
  dementia ~ age + depression_history + education_years + physical_activity +
    diabetes + hypertension + BMI + I(BMI^2),
  data = synthetic_data,
  family = binomial()
)
summary(m2)
exp(coef(m2))
# check model 
AIC(m1, m2)
```

# Conclusion
From the results, it can be seen that in both models, the risk of dementia increases with age and depression, while the risks of diabetes and hypertension increase. Adding the secondary BMI term significantly improves the model's fit and confirms the expected non-linear (U-shaped) relationship between BMI and the risk of dementia.